### 2016-10-25 writing/paper/sections/introduction.tex
mesh patterns, minor changes to introduction

The study of permutation patterns began as a result of Knuth's statements on stack sorting in
\emph{The Art of Computer Programming}\cite[p.~243, Ex.~5,6]{Knuth:1997:ACP:260999}. This original
concept---a subsequence of symbols having a particular relative order, now known as classical
patterns---has been expanded to a variety of definitions. Mesh patterns \textcite{babstein2000} 
considered \emph{vincular} patterns---also known as \emph{generalised} or \emph{dashed} 
patterns---where two adjacent entries in the pattern must also be adjacent in the permutation. 
\textcite{MR2652101} look at classes of pattern where both columns and rows can be shaded, these are
called \emph{bivincular} patterns. \emph{Bruhat-restricted} patterns were studied by 
\textcite{MR2264071} in order to establish necessary conditions for a Schubert variety to be
Gorenstein. All of these definitions are subsumed under the definition of \emph{mesh patterns},
introduced by \textcite{journals/combinatorics/BrandenC11} to capture explicit expansions for 
certain permutation statistics.

>>>>

The study of permutation patterns began as a result of Knuth's statements on stack sorting in
\emph{The Art of Computer Programming}\cite[p.~243, Ex.~5,6]{Knuth:1997:ACP:260999}. This original
concept---a subsequence of symbols having a particular relative order, now known as classical
patterns---has been expanded to a variety of definitions. Mesh patterns \textcite{babstein2000} 
considered \emph{vincular} patterns (also known as \emph{generalised} or \emph{dashed} 
patterns) where two adjacent entries in the pattern must also be adjacent in the permutation. 
\textcite{MR2652101} look at classes of pattern where both columns and rows can be shaded, these are
called \emph{bivincular} patterns. \emph{Bruhat-restricted} patterns were studied by 
\textcite{MR2264071} in order to establish necessary conditions for a Schubert variety to be
Gorenstein. All of these definitions are subsumed under the definition of \emph{mesh patterns},
introduced by \textcite{journals/combinatorics/BrandenC11} to capture explicit expansions for 
certain permutation statistics.

### 2016-10-25 writing/paper/sections/meshpatterns.tex
mesh patterns, minor changes to introduction

\chapter{Introduction\label{cha:introduction}}
\section{What is a Permutation?}
In \emph{The Art of Computer Programming} \cite[p.~45]{Knuth:1997:ACP:260999} Knuth states,
\enquote{A \emph{permutation of \(n\) objects} is an arrangement of \(n\) distinct
objects in a row}.
When considering permutations we can consider them as
occurring on the set \(\nrange{n} = \setrange{1}{n}\), therefore a permutation
is a \emph{bijection} \(\pi:\nrange{n} \mapsto \nrange{n}\). %A
% permutation \(\pi\) can be written in two-line notation
% \begin{equation*}
% \pi = \twoline{1,2,\dotso,n}{\pi(1),\pi(2),\dotso,\pi(n)}
% \end{equation*}
% However,
This
% most frequent
notation
% used in computer science
is called
\emph{one-line notation}.
% in this form we drop the top line of the two line
% notation and are left with the following.
In this form we write the entries of the permutation in order, and get
\begin{equation*}
\pi = \perm{\pi(1),\pi(2),\dotso,\pi(n)}
\end{equation*}

\begin{example}
The 6 permutations on \(\nrange{3}\), in one-line notation, are
\begin{equation*}
\perm{1,2,3},\perm{1,3,2},\perm{2,1,3},\perm{2,3,1}, \perm{3,1,2}, \perm{3,2,1}
\end{equation*}
\end{example}

We can display a permutation in a \emph{plot} in order to give a graphical
representation of the permutation. In such a plot we display the points
\((i,\pi(i))\) in a Cartesian coordinate system. The plot of the permutation
\(\pi = \perm{2,3,1}\) is shown below
\begin{equation*}
    \pattern{}{2,3,1}{}
\end{equation*}
It is convenient to call the elements of the permutation \emph{points} when
referring to these plots.

The set of all permutations of length \(n\) is \(\mathfrak{S}_n\) and
has size \(n!\). The set of all permutations is
\(\mathfrak{S}=\bigcup_{i=0}^{\infty}\mathfrak{S}_i\). Note that \(\mathfrak{S}_0\)
has exactly one element, the empty permutation \(\varepsilon\). As a function
this is equivalent to the unique bijection \(\emptyset \mapsto \emptyset\), and
it's one-line representation is the empty string.

\section{Classical Permutation Patterns}
Classical permutation patterns began to be studied as a result of Knuth's
statements about stack-sorting in \emph{The Art of Computer Programming}
\cite[p.~243, Ex.~5,6]{Knuth:1997:ACP:260999}.

@@

A permutation \(\pi\) is said to \emph{contain} the permutation \(\sigma\) of
length \(k\) as a \emph{pattern} (denoted \(\sigma \preceq \pi\)) if there is some
subsequence \(i_1i_2\dotsm{}i_k\) such that the sequence
\(\pi(i_1)\pi(i_2)\dotsm\pi(i_k)\) is order isomorphic to
\(\sigma(1)\sigma(2)\dotsm\sigma(k)\). If \(\pi\) does not contain
\(\sigma\), we say that \(\pi\) \emph{avoids} \(\sigma\).

For example the permutation \(\pi = \perm{2,4,1,5,3}\) contains the pattern
\(\sigma = \perm{2,3,1}\), since the second, fourth and fifth elements
(\(453\)) are order isomorphic to \(\perm{231}\).
This can be seen graphically below, the subsequence order isomorphic to \(\sigma\)
is highlighted.
\begin{equation*}
    \pattern{2,4,5}{2,4,1,5,3}{}
\end{equation*}

We denote the set of permutations of length \(n\) avoiding a pattern \(\sigma\)
as \(\Av_n(\sigma)\) and \(\av{\sigma} = \bigcup_{i=0}^{\infty}\Av_i(\sigma)\).

Knuth's statements were exercises in showing that the permutations avoiding the
pattern \(\perm{2,3,1}\) are precisely the permutations that are sortable to
the identity permutation using a single stack, and that permutations avoiding the
pattern \(\perm{3,2,1}\) are precisely the permutations that are sortable to
the identity permutation using a single queue with bypass.

\section{Mesh Patterns}
When looking at a classical pattern, for example \(\perm{2,3,1}\), any occurrences can be arbitrarily
placed. However it may be interesting to consider occurrences where the elements
corresponding to the \(2\) and \(3\) are adjacent in the permutation. In the past
these sort of questions have lead to a variety of definitions.  \textcite{babstein2000}
considered \emph{vincular} patterns (also known as \emph{generalised} or \emph{dashed} patterns),
those where two adjacent entries in the pattern must be adjacent in the permutation.
\textcite{MR2652101} look at classes of pattern where both columns and rows
can be shaded, these are called \emph{bivincular} patterns.
\emph{Bruhat-restricted} patterns were studied by \textcite{MR2264071} in order
to establish necessary conditions for a Schubert variety to be Gorenstein.
Mesh patterns also encompass a subset of \emph{barred} patterns introduced by
\textcite{MR2716312}, those with only one barred letter.

All of these definitions are subsumed under the definition of \emph{mesh patterns},
introduced by \textcite{journals/combinatorics/BrandenC11} to
capture explicit expansions for certain permutation statistics. They are a
natural extension of classical permutation patterns.
A \emph{mesh pattern} is a pair
\begin{equation*}
    p = (\tau,R)\text{ with } \tau \in \mathfrak{S}_k \text{ and } R \subseteq
    [0,k]\times [0,k].
\end{equation*}

The set \(R\) is called the \emph{mesh set} of the mesh pattern \(p\).
The plot for a mesh pattern looks similar to that of a classical pattern with the
addition that we shade the unit square with bottom left corner \((i,j)\) for each \((i,j) \in R\):
\begin{equation*}
    \pattern{}{ 2, 1, 3 }{ 0/1, 0/2, 1/0, 1/1, 2/1, 2/2 }
\end{equation*}

The example of an occurrence of \(\perm{2,3,1}\) with adjacent elements \(23\)
can be represented as a mesh pattern by
\begin{equation*}
  \pattern{}{2,3,1}{1/0,1/1,1/2,1/3}
\end{equation*}

We define containment (denoted \(p\preceq \pi\)), and avoidance, of the pattern \(p\) in the permutation
\(\tau\) on mesh patterns analogously to classical containment, and avoidance,
of \(\pi\) in \(\tau\) with the additional restrictions on the relative
position of the occurrence of \(\pi\) in \(\tau\). These restrictions say that
no elements of \(\tau\) are allowed in the regions of the plot corresponding
to shaded boxes in the mesh.

@@

Now we consider the plot for the permutation \(\sigma\). This permutation
avoids \(p\) since for every occurrence of the classical pattern \(\perm{2,1,3}\)
there is at least one point in one of the shaded boxes. For example, consider the
subsequence \(315\) in \(\sigma\), this is an occurrence of \(\perm{2,1,3}\) but
not the mesh pattern since the points with values \(4\) and \(2\) are in the
regions corresponding to the boxes \((0,1)\) and \((0,2)\), which are shaded in
\(p\). This is shown in the figure below.

@@

We denote the avoidance sets for mesh patterns in the same way as for
classical patterns. Given a mesh pattern \(p=(\sigma,R)\) we say that \(\sigma\) is the \emph{underlying
classical pattern} of \(p\).
\begin{note}
    \label{not:classmesh}
    Classical patterns can be thought of as a subset of mesh patterns: the
    classical pattern \(\pi\) can be represented by a mesh pattern as
    \((\pi,\emptyset)\).
\end{note}
\begin{note}
  Permutations can also be thought of as a special case of mesh patterns with
  the mesh set being \([0,k]\times [0,k]\)
\end{note}

Two patterns are said to be \emph{coincident} if they are avoided by the same set of
permutations and \emph{Wilf-equivalent} if they are avoided by the same number of permutations
at every length.

Avoiding pairs of patterns of the same length with certain properties has
also been studied in the past, \textcite{MR2178749} considered avoiding a
pair of vincular patterns of length 3. \textcite{2015arXiv151203226B} study avoiding
a vincular and a covincular pattern simultaneously in order to achieve some
interesting counting results. However, very little work has been done on avoiding
a mesh pattern and a classical pattern simultaneously. In this work we aim to
establish some ground in this field by computing coincidences and Wilf-classes
and calculating some of the enumerations of avoiders of a mesh pattern of length
2 and a classical pattern of length 3.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "msc-tannock-2016"
%%% End:

>>>>

A \emph{permutation} is a bijection from the set \(\nrange{n} = \setrange{1}{n}\) to itself. The
set of all such bijections on this set if denoted \(\mathfrak{S}_n\) and has size \(n!\). We can
denote an individual permutation \(\pi\in\mathfrak{S}_n\) in \emph{one-line notation} by writing the
entries of the permutation in order, therefore \(\pi = \perm{\pi(1),\pi(2),\dotso,\pi(n)}\). The
set \(\mathfrak{S}_n\) has exactly one element, the empty permutation \(\varepsilon\).

@@

The definition of order isomorphism allows us to give the meaning of containment for classical
permutation patterns.
\begin{definition}
    A permutation \(\pi\in\mathfrak{S}_n\) contains the pattern \(\sigma\in\mathfrak{S}_k\) (\mbox{denoted} 
    \(\sigma \preceq \pi\)) if there is some subsequence of indices of \(\pi, i_1i_2\dotsm{}i_k\)
    such that the sequence \(\pi(i_1)\pi(i_2)\dotsm\pi(i_k)\) is order isomorphic to
    \(\sigma(1)\sigma(2)\dotsm\sigma(k)\). If \(\pi\) does not contain \(\sigma\), we say that 
    \(\pi\) \emph{avoids} \(\sigma\).
\end{definition}
\begin{example}
\label{ex:contexmpl}
The permutation \(\pi = \perm{2,4,1,5,3}\) contains the pattern 
\(\sigma = \perm{2,3,1}\), since the second, fourth and fifth elements (\(453\)) are order 
isomorphic to \(\perm{231}\), it also contains the occurrence \(241\). The permutation 
\(\perm{2,4,1,5,3}\) avoids the pattern \(\perm{3,2,1}\).
\end{example}

We denote the set of permutations of length \(n\) avoiding a pattern \(\sigma\) as \(\Av_n(\sigma)\)
and \(\av{\sigma}=\bigcup_{i=0}^{\infty}\Av_i(\sigma)\).

We can display a permutation graphically in a \emph{plot}, in such a plot we display the points
\(G(\pi) = \{(i,\pi(i))\mid i\in[1,n]\}\) in a Cartesian coordinate system. The plots of the permutations \(\pi = 
\perm{2,4,1,5,3}\) and \(\sigma = \perm{2,3,1}\) can be seen in \FigureRef{fig:plots}.
\FigureRef{fig:containment} shows the containment of \(\sigma\) in \(\pi\) as in \ExampleRef{ex:contexmpl}.

\begin{figure}[htb]
    \begin{center}
    \raisebox{6ex}{\(G(\pi)=\)}
    \begin{tikzpicture}[scale=\picscale]
        \modpattern{}{2,4,1,5,3}{}
    \end{tikzpicture}
    \raisebox{6ex}{\(\quad{}G(\sigma)=\)}
    \raisebox{0.35cm}{
    \begin{tikzpicture}[scale=\picscale]
        \modpattern{}{2,3,1}{}
    \end{tikzpicture}}

        \caption{The plots of the permutations \(\pi\) and \(\sigma\).}
        \label{fig:plots}
    \end{center}
\end{figure}

\begin{figure}[htb]
    \begin{center}
    \(\pattern{2,4,5}{2,4,1,5,3}{}\)
        \caption{The occurrence of 231 in 24153 corresponding to 453.}
        \label{fig:containment}
    \end{center}
\end{figure}

\begin{definition}
A \emph{mesh pattern} is a pair
\begin{equation*}
    p = (\tau,R)\text{ with } \tau \in \mathfrak{S}_k \text{ and } R \subseteq [0,k]\times [0,k].
\end{equation*}
\end{definition}
We define containment (denoted \(p\preceq \pi\)), and avoidance, of the pattern \(p\) in the permutation
\(\tau\) on mesh patterns analogously to classical containment, and avoidance,
of \(\pi\) in \(\tau\) with the additional restrictions on the relative
position of the occurrence of \(\pi\) in \(\tau\). These restrictions say that
no elements of \(\tau\) are allowed in the regions of the plot corresponding
to shaded boxes in the mesh.

@@

Now we consider the plot for the permutation \(\sigma\). This permutation
avoids \(p\) since for every occurrence of the classical pattern \(\perm{2,1,3}\)
there is at least one point in one of the shaded boxes. For example, consider the
subsequence \(315\) in \(\sigma\), this is an occurrence of \(\perm{2,1,3}\) but
not the mesh pattern since the points with values \(4\) and \(2\) are in the
regions corresponding to the boxes \((0,1)\) and \((0,2)\), which are shaded in
\(p\). This is shown in the plot below.

@@

We denote the avoidance sets for mesh patterns in the same way as for
classical patterns. Given a mesh pattern \(p=(\sigma,R)\) we say that \(\sigma\) is the \emph{underlying
classical pattern} of \(p\).
